--- 
title: 'Lecture 3:  Data Manipulation'
date: "October 10, 2017" 
output: 
  revealjs::revealjs_presentation: 
    self_contained: false
    lib_dir: libs
    css: class_big_fonts.css
    highlight: default 
    smart: true
    center: true
    transition: default 
    background_transition: default 
    text-align: left;
    reveal_options: 
      slideNumber: true 
      previewLinks: true 
---

```{r global_options, echo = FALSE, include = FALSE}
options(width = 80)
knitr::opts_chunk$set(cache = FALSE, tidy = FALSE, size = "small")
library(dplyr)
```


## Contents

* `apply` family functions
* Intro to `dplyr` package
* Data manipulation functions
* Chaining operations
* Joining & appending datasets
* Extras

# apply family

## apply, lapply, sapply functions

- The `apply` family functions, are 
**functions which manipulate slices of data**
stored as matrices, arrays, lists and data-frames 
**in a repetitive way**. 

- These functions **avoid the explicit use of loops**, and might 
be **more computationally efficient**, depending on how big a dataset is. 
For more details on runtimes see this [link](https://www.r-bloggers.com/gnu-r-loop-speed-comparison/).

- `apply` allow you to perform operations with 
**very few lines of code.**

- The family comprises: **apply, lapply , sapply, vapply, 
mapply, rapply, and tapply**. The difference lies in the structure
of input data and the desired format of the output).


## apply function

`apply` operates on arrays/matrices. 

In the example below we obtain column sums of matrix `X`.

```{r}
(X <- matrix(sample(30), nrow = 5, ncol = 6))
apply(X, MARGIN = 2 , FUN = sum)
```

**Note:** that in a matrix `MARGIN = 1` indicates rows and `MARGIN = 2`
indicates columns.


## apply function


<div class="left", style="width: 50%">
- `apply` can be used with **user-defined functions**:

```{r}
print(X)

# number entries < 15
apply(X, 2, function(x) 10*x + 2) 
```

</div>

<div class="right", style="width: 50%">
- The function can be defined outside `apply()`,


```{r}
logColMeans <- function(x, eps = NULL) {
  if (!is.null(eps)) x <- x + eps
  return(mean(x))
}
apply(X, 2, logColMeans) 
apply(X, 2, logColMeans, eps = 0.1) 
```

</div>

## lapply/sapply functions

<div class="left", style="width: 50%">

- `lapply()` is used to **repeatedly apply a function to 
elements of a sequential object** such as a vector, list, 
or data-frame (applies to columns).

- The **output is a list** with the same number of 
elements as the input object.

- `sapply` is the same as `lapply` but **returns a "simplified" output**.

- user-defined functions can be used with `sapply/lapply`

</div>

<div class="right", style="width: 50%">

```{r}
# lapply returns a list
lapply(1:3, function(x) x^2)
# which you can 'simplify' with unlist()
unlist(lapply(1:3, function(x) x^2))
# Or you could use sapply() instead
sapply(1:3, function(x) x^2)
```

</div>

 
## mapply functions


- `mapply` stands for 'multivariate' apply.  It **applies a function
to a multiple list or multiple vectors as arguments**. 
- The goal is to vectorize arguments to a function which usually 
does not accept vectors as arguments. 

```{r}
# function word() returns a string of character C repeated k times.
word <- function(C,k) paste(rep.int(C,k), collapse='')
mapply(word, LETTERS[1:6], 6:1, SIMPLIFY = FALSE)
```


## Exercise 1

</br>

- Go to the "Lec3_Exercises.Rmd" file, which can be downloaded
from the class website under the Lecture tab.

- Complete Exercise 1.


# `dplyr` package


## `dplyr`

* Introduces **a grammar of data manipulation**.
* **Code-efficient** for data exploration and transformation.
* **Fast on data frames** (written in C++): has speed of C and ease of R.
* **Intuitive to write and easy to read**, esp. when using the *chaining* syntax.
* You should use `dplyr` even as a beginner R user, and [here is why](http://varianceexplained.org/r/teach-tidyverse/).

```{r, eval=FALSE}
# To install dplyr with latest updates
install.packages("devtools")
devtools::install_github("tidyverse/dplyr")

# Or you could use CRAN
install.packages("dplyr")
```


## tibbles

> Tibbles are a modern take on data frames. They keep the features that have 
stood the test of time, and drop the features that used to be convenient 
but are now frustrating (i.e. converting character vectors to factors).


* A tibble, `tbl`, is **a wrapper for a data frame that prints nicely**.
* The print method for `tbl` shows only the **first 10 rows, 
and all the columns that fit on screen**. 
* Each column is also reported together its type.
* Tibbles (and `dplyr`) **do NOT preserve the row names**. 
* **Subsetting `tbl` is strickter** than subsetting `data.frames`,
and ALWAYS returns objects with expected class, i.e. with a single
`[` you get back a tibble, and with double`[[` you get a vector.



## Movie industry dataset

`movies.csv` contains information on last three decades of movies. 
The data has been scraped from the IMDb website and can be accessed from a [github repo](https://raw.githubusercontent.com/Juanets/movie-stats/master/movies.csv).


```{r, width = 100}
url <- "https://raw.githubusercontent.com/Juanets/movie-stats/master/movies.csv"
movies.df <- read.csv(url)
rownames(movies.df) <- paste0("M", 1:nrow(movies.df))
dim(movies.df)

colnames(movies.df)
```

##

```{r}
head(movies.df)
```


## Convert to tibble

```{r}
library(dplyr)

# convert to tibble
movies <- tbl_df(movies.df)
class(movies)

# printing only shows 10 rows and as many columns as can fit on your screen
movies
```

<small> **Note:** Tibbles and dplyr **do NOT preserve the row names**, you need
to create an "id" column instead. </small>



## 
```{r}
str(movies)
```


# Data manipulation functions

## `dplyr` verbs

The most commonly used `dplyr` functions (or basic verbs) are:

* `filter()`: keep rows matching criteria, 
* `select()`: pick columns by name, 
* `arrange()`: reorder rows, 
* `mutate()`: add new variables, 
* `summarise()`: reduce variables to values

</br>

Operations performed with the above functions can be done using **base R 
functions, but they would be less computationally efficient, and 
require writing more lines of (ugly) code.** 

</br>

Learn about `dplyr` from the [turtorial written by its creator, Hadley Wickham](https://www.dropbox.com/sh/i8qnluwmuieicxc/AAAgt9tIKoIm7WZKIyK25lh6a?preview=dplyr-tutorial.pdf)



## Structure of `dplyr` functions

* the first argument is a data frame 
* subsequent argument specify what to do
* always return a data frame



## `filter()`: keep rows matching criteria

```{r echo = FALSE}
library(hflights)
flights <- tbl_df(hflights)
```


```{r eval = FALSE}
# base R approach to find all comedies by Woody Allen
movies[movies$genre == "Comedy" & movies$director == "Woody Allen", ]
```

```{r}
# dplyr approach
# note: both comma or ampersand represent AND condition
filter(movies, genre == "Comedy", director == "Woody Allen")
```

##

```{r}
# use pipe for OR condition
filter(movies, country == "Greece" | country == "Chile")
```

## 

```{r}
# you can also use %in% operator
print(filter(movies, country %in% c("Argentina", "Colombia", "Chile")), n = Inf)
```


## `select()`: pick columns by name


```{r, eval = FALSE}
# base R approach to select columns
movies[, c("name", "year", "genre")]
```

```{r}
# dplyr approach
movies.sub <- select(movies, name, country, year, genre)
movies.sub
```

##

```{r}
# use colon to select multiple contiguous columns, 
select(movies, name, genre:score)
```



## `select()` helpers

You can use the following functions to help select the columns:

* `starts_with()`
* `ends_with()`
* `contains()`
* `matches()` (matches a regular expression)

```{r}
select(movies, starts_with("r"))
```

## 

```{r}
select(movies, ends_with("e"))
```


```{r}
select(movies, contains("re"))
```


## Dropping columns

```{r}
# remove budget and company columns
print(select(movies, -budget, -company), n = 6)

# Selecting and renaming in one
print(select(movies, name, gross_revenue = gross), n = 6)
```


## `arrange()`: reorder rows


```{r}
# dplyr approach
print(arrange(movies.sub, name), n = 6)

# use `desc` for descending
print(arrange(movies.sub, desc(year)), n = 6)
```

## `mutate()`: add new variables


```{r eval = FALSE}
# base R approach to create a new variable 'profit'
movies$profit <- movies$gross - movies$budget
```

```{r}
# dplyr approach
movies <- mutate(movies, profit = gross - budget)
select(movies, name, gross, budget, profit)
```

##

Generating multiple new variables

```{r}
movies <- mutate(movies, 
                 profit = gross - budget, 
                 gross_in_mil = gross/10^6,
                 budget_in_mil = budget/10^6,
                 profit_in_mil = profit/10^6)
select(movies, name, contains("_in_mil"))
```

## `summarise()`: reduce variables to values


<div class="left", style="width: 50%">

* `summarize()` can be used to aggregate data or to compute
a summarizing value of interest.

```{r}
# Total gross revenue for all movies in 30 years:
summarise(movies, 
          tot_gross_in_bil = sum(gross)/10^9)
```


</div>

<div class="right", style="width: 50%">

* `summarize()` is **primarily useful on data previously grouped by 
one or more variables** using `group_by()`.

```{r}
by_genre <- group_by(movies, genre)
summarize(by_genre, total = sum(gross)/10^9)
```

</div>


## Useful summary functions

* `min(x)`, `median(x)`,`max(x)`, `quantile(x, p)`
* `n()`, `n_distinct()`, `sum(x)`, `mean(x)`
* `sum(x > 10)`, `mean(x > 0)`
* `sd(x)`, `var(x)`

# Chaining operations

## Pipe operator:  `%>%`

<div class="left", style="width: 50%">

* The `%>%`  operator was intoduced in `magrittr` package to 
**pipe values forward into an expression or function call**.
* In the pipe notation we have something like `x %>% f(y)`, rather than `f(x, y)`
* This is similar to the [Unix pipes](http://www.linfo.org/pipe.html): 
`|` used to **send the output of one program to another program for 
further processing**.
</div>

<div class="right", style="width: 50%">


```{r fig.width=6, fig.height=4.5}
rnorm(500) %>%
  matrix(ncol = 50) %>%
  colSums() %>%
  plot(x = 1:50)
```

</div>

## Chaining operations

* Pipe operators used together with `dplyr` functions make a large
difference as they semantically change your code in a way that 
**makes it more intuitive to both read and write**. 

* The pipes allow users to chain operators which reflects the 
**sequential nature of data-processing tasks**.

* **Chaining increases readability** significantly when there are 
many commands 

* `%>%` operator is automatically imported into `dplyr`

##

<small> Find movies from USA produced after 2010. Group by genre and
compute the group mean gross revenue in million dollars. Then print the genre
mean 'gross' revenue arranged in a descending order: </small>


<div class="left", style="width: 50%">
```{r}
# nesting 
arrange(
  summarise(
    group_by(
      filter(movies, 
             year > 2010, country == "USA"
      ),
      genre
    ),
    mean_gross = mean(gross)/10^6
  ),
  mean_gross
)
```


</div>

<div class="right", style="width: 50%">

```{r}
# chaining 
movies %>%
  filter(year > 2010, country == "USA") %>%
  group_by(genre) %>%
  summarise(mean_gross = mean(gross)/10^6) %>%
  arrange(mean_gross)
```

</div>

## Exercises 2


</br>

- Go to the "Lec3_Exercises.Rmd" file, which can be downloaded
from the class website under the Lecture tab.

- Complete Exercise 2.


# Joining & appending datasets

## Merging two tables



<div class="left", style="width: 50%">
```{r}
x <- data.frame(
  item = c("cherries", "orange", 
           "broccoli", "blueberries", 
           "cabbage", "lemon", "spinach", 
           "eggplants"),
  color = c("red", "orange", "green", 
            "blue", "green", "yellow", 
            "green", "purple"),
  stringsAsFactors = FALSE)

y <-  data.frame(
  item = c("carrot", "cherries", 
           "broccoli", "blueberries", 
           "cabbage", "lemon", "tomato"),
  class = c("vegetable", "fruit", 
            "vegetable", "fruit", 
            "vegetable", "fruit", "fruit"), 
  stringsAsFactors = FALSE)
```

</div>

<div class="right", style="width: 50%">


```{r, echo = FALSE}
library(kableExtra); library(knitr)
kable(x, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "royalblue")

kable(y, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "darkolivegreen")
```

</div>


## `inner_join(x, y)`

* Include only rows in both x and y

```{r, echo=TRUE, message=TRUE, warning=TRUE, results="hide"}
inner_join(x, y) # or x %>% inner_join(y)
```


```{r, echo = FALSE, message=FALSE, warning=FALSE}
kable(x, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "royalblue") %>%
  add_header_above(c("x" = 2))

kable(y, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "darkolivegreen") %>%
  add_header_above(c("y" = 2))

kable(inner_join(x, y), "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_right") %>%
  column_spec(1:3, background = "bisque") %>%
  add_header_above(c("inner_join(x, y)" = 3))
```


## `left_join(x, y)`: 

* Include all of x, and matching rows of y

```{r, echo=TRUE, message=TRUE, warning=TRUE, results="hide"}
left_join(x, y) # or x %>% left_join(y)
```


```{r, echo = FALSE, message=FALSE, warning=FALSE}
kable(x, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "royalblue") %>%
  add_header_above(c("x" = 2))

kable(y, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "darkolivegreen") %>%
  add_header_above(c("y" = 2))

kable(left_join(x, y), "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_right") %>%
  column_spec(1:3, background = "bisque") %>%
  add_header_above(c("left_join(x, y)" = 3))
```


## `semi_join(x, y)`: 

* Include rows of x that match y

```{r, echo=TRUE, message=TRUE, warning=TRUE, results="hide"}
semi_join(x, y) # or x %>% semi_join(y)
```

```{r, echo = FALSE, message=FALSE, warning=FALSE}
kable(x, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "royalblue") %>%
  add_header_above(c("x" = 2))

kable(y, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "darkolivegreen") %>%
  add_header_above(c("y" = 2))

kable(semi_join(x, y), "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_right") %>%
  column_spec(1:2, background = "bisque") %>%
  add_header_above(c("semi_join(x, y)" = 2))
```


## `anti_join(x, y)`: 

* Include rows of x that don’t match y

```{r, echo=TRUE, message=TRUE, warning=TRUE, results="hide"}
anti_join(x, y) # or x %>% anti_join(y)
```

```{r, echo = FALSE, message=FALSE, warning=FALSE}
kable(x, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "royalblue") %>%
  add_header_above(c("x" = 2))

kable(y, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "darkolivegreen") %>%
  add_header_above(c("y" = 2))

kable(anti_join(x, y), "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_right") %>%
  column_spec(1:2, background = "bisque") %>%
  add_header_above(c("anti_join(x, y)" = 2))
```


# Extra functions

## Renaming

```{r}
movies.sub <- movies %>% select(name, director, year, score, gross)
print(movies.sub, n = 3)
# Renaming variable
movies.sub %>% rename(gross_revenue = gross)
```

## Distinct values

```{r}
# Unique values
movies %>% distinct(rating)
```

**Note that** can take on multiple variables, and would return distinct variable combinations.

## Group counts

Use a `tally()` function to generate a group frequency table:

```{r}
movies %>% group_by(genre) %>% tally()
```


## Window Functions

* Aggregation functions such as `mean()`, `n()` return 1 value per group.
* **Window functions return multiple values per group**. Examples include:
ranking and ordering functions (like `min_rank`, `top_n()`), 
offset functions (`lead` and `lag`), and cumulative aggregates (like `cummean`).



```{r}
# rewrite more simply with the `top_n` function
movies %>%
  select(name, genre, year, score) %>%
  group_by(genre) %>%  
  top_n(2, wt = score) %>%      # if 'wt' argument in top_n() is not specified
  arrange(genre, year, score)   # the last variable is taken for ordering
```  



## Sampling functions

```{r}
# randomly sample a fixed number of rows, without replacement
movies %>% sample_n(15)
```

##

```{r}
# randomly sample a fraction of rows, with replacement
movies %>% sample_frac(0.01, replace=TRUE)
```

