---
title: "Left over slides"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## When should you use R?

- R can be used for **data analysis and visualization**. 

> - It provides efficient **exploration and prototyping** tools, 
but also offers powerful frameworks for **building and fitting statistical models**.

> - R is very good when working with **data that fits in the computer’s memory**.

> - It’s NOT as good for working with massive datasets whose elements 
need to be continuously accessed.



## Data import 

- **Excel files**. To read the excel files you need
to first install a package "xlsx". Simply run 
`install.packages("xlsx")`.

```{r eval = FALSE}
# first row contains variable names
library(xlsx)
# read the 1st worksheet
mydata <- read.xlsx("/path/to/filename.xlsx", 1)  

# read in the worksheet named mysheet
mydata <- read.xlsx("/path/to/filename.xlsx", sheetName = "mysheet")
```

- **SAS files**. You need to install the `Hmisc` or `foreign` 
packages first.

```{r eval = FALSE}
# First, save your SAS dataset in trasport (xport) format
libname out xport 'path/to/filename.xpt';
data out.mydata;
set sasuser.mydata;
run;

# Then, in R 
library(Hmisc)
mydata <- sasxport.get("path/to/filename.xpt")
# character variables are converted to R factors
# or
library(foreign)
mydata <- read.xport("path/to/filename.xpt")
```



## Exporting Data

- **Text files**:
```{r eval = FALSE}
write.table(mydata, "path/to/filename.txt", sep="\t")  # tab-delimited
write.table(mydata, "path/to/filename.csv", sep=",")   # comma-delimited
write.csv(mydata, "path/to/filename.csv")              # comma-delimited
```

- **Excel spreadsheet**:

```{r eval = FALSE}
library(xlsx)
write.xlsx(mydata, "path/to/filename.xlsx")
```

- **SAS** files:
```{r eval = FALSE}
# write out text datafile and an SAS program to read it
library(foreign)
write.foreign(mydata, datafile = "path/to/filename.txt", 
              codefile = "path/to/filename.sas",  package="SAS")
```

Here are links to more details on data 
[import](http://www.statmethods.net/input/importingdata.html)
and [export](http://www.statmethods.net/input/exportingdata.html). 



## Identifiers 

**Google Convention:** 
  
  * No underscores ( _ ) or hyphens ( - ) in identifiers. 
* **Variable names** all with lower case letters; words separated with dots 
('variable.name'), but 'variableName' is also accepte
* **Function names** have initial capital letters and no dots 
('FunctionName'). Function names should be verbs. Constants are named like 
functions but with an initial k.

```{r, eval = FALSE}
# Good
avg.clicks            # variable name
CalculateAvgClicks    # function name
# Bad
avg_Clicks                                 # variable name
calculate_avg_clicks , calculateAvgClicks  # function name
```

## Identifiers 

**Hadley Convention:**
  
  - Variable and function names should be lowercase. 
- Use an underscore (_) to separate words within a name. 
- Generally, variable names should be nouns and function names should be verbs. 
- Use names that are concise and meaningful. 

```{r eval = FALSE}
# Good
day_one
day_1

# Bad
first_day_of_the_month
DayOne
dayone
djm1
```


## Spacing

- Place spaces around all infix operators (=, +, -, <-, etc.). 

- Always put a space after a comma, and never before (just like in regular English).

- Place a space before left parentheses, except in a function call.

```{r eval = FALSE}
# Good
average <- mean(feet / 12 + inches, na.rm = TRUE)
if (debug) do(x)
plot(x, y)
if (debug) do(x)
diamonds[5, ]

# Bad
average<-mean(feet/12+inches,na.rm=TRUE)
if(debug)do(x)
plot (x, y)
if ( debug ) do(x)  # No spaces around debug
x[1,]               # Needs a space after the comma
x[1 ,]              # Space goes after comma not before
```





```{r}
sessionInfo()
```

