---
title: "Left over slides"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## When should you use R?

- R can be used for **data analysis and visualization**. 

> - It provides efficient **exploration and prototyping** tools, 
but also offers powerful frameworks for **building and fitting statistical models**.

> - R is very good when working with **data that fits in the computer’s memory**.

> - It’s NOT as good for working with massive datasets whose elements 
need to be continuously accessed.


# Joining & appending datasets

## Merging two tables

<div class="left", style="width: 50%">
```{r}
x <- data.frame(
  item = c("cherries", "orange", 
           "broccoli", "blueberries", 
           "cabbage", "lemon", "spinach", 
           "eggplants"),
  color = c("red", "orange", "green", 
            "blue", "green", "yellow", 
            "green", "purple"),
  stringsAsFactors = FALSE)

y <-  data.frame(
  item = c("carrot", "cherries", 
           "broccoli", "blueberries", 
           "cabbage", "lemon", "tomato"),
  class = c("vegetable", "fruit", 
            "vegetable", "fruit", 
            "vegetable", "fruit", "fruit"), 
  stringsAsFactors = FALSE)
```

</div>

<div class="right", style="width: 50%">


```{r, echo = FALSE}
library(kableExtra); library(knitr)
kable(x, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "royalblue")

kable(y, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "darkolivegreen")
```

</div>


## `inner_join(x, y)`

* Include only rows in both x and y

```{r, echo=TRUE, message=TRUE, warning=TRUE, results="hide"}
inner_join(x, y) # or x %>% inner_join(y)
```


```{r, echo = FALSE, message=FALSE, warning=FALSE}
kable(x, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "royalblue") %>%
  add_header_above(c("x" = 2))

kable(y, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "darkolivegreen") %>%
  add_header_above(c("y" = 2))

kable(inner_join(x, y), "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_right") %>%
  column_spec(1:3, background = "bisque") %>%
  add_header_above(c("inner_join(x, y)" = 3))
```


## `left_join(x, y)`: 

* Include all of x, and matching rows of y

```{r, echo=TRUE, message=TRUE, warning=TRUE, results="hide"}
left_join(x, y) # or x %>% left_join(y)
```


```{r, echo = FALSE, message=FALSE, warning=FALSE}
kable(x, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "royalblue") %>%
  add_header_above(c("x" = 2))

kable(y, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "darkolivegreen") %>%
  add_header_above(c("y" = 2))

kable(left_join(x, y), "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_right") %>%
  column_spec(1:3, background = "bisque") %>%
  add_header_above(c("left_join(x, y)" = 3))
```


## `semi_join(x, y)`: 

* Include rows of x that match y

```{r, echo=TRUE, message=TRUE, warning=TRUE, results="hide"}
semi_join(x, y) # or x %>% semi_join(y)
```

```{r, echo = FALSE, message=FALSE, warning=FALSE}
kable(x, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "royalblue") %>%
  add_header_above(c("x" = 2))

kable(y, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "darkolivegreen") %>%
  add_header_above(c("y" = 2))

kable(semi_join(x, y), "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_right") %>%
  column_spec(1:2, background = "bisque") %>%
  add_header_above(c("semi_join(x, y)" = 2))
```


## `anti_join(x, y)`: 

* Include rows of x that don’t match y

```{r, echo=TRUE, message=TRUE, warning=TRUE, results="hide"}
anti_join(x, y) # or x %>% anti_join(y)
```

```{r, echo = FALSE, message=FALSE, warning=FALSE}
kable(x, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "royalblue") %>%
  add_header_above(c("x" = 2))

kable(y, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_left") %>%
  column_spec(1:2, background = "darkolivegreen") %>%
  add_header_above(c("y" = 2))

kable(anti_join(x, y), "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_right") %>%
  column_spec(1:2, background = "bisque") %>%
  add_header_above(c("anti_join(x, y)" = 2))
```


## Data import 

- **Excel files**. To read the excel files you need
to first install a package "xlsx". Simply run 
`install.packages("xlsx")`.

```{r eval = FALSE}
# first row contains variable names
library(xlsx)
# read the 1st worksheet
mydata <- read.xlsx("/path/to/filename.xlsx", 1)  

# read in the worksheet named mysheet
mydata <- read.xlsx("/path/to/filename.xlsx", sheetName = "mysheet")
```

- **SAS files**. You need to install the `Hmisc` or `foreign` 
packages first.

```{r eval = FALSE}
# First, save your SAS dataset in trasport (xport) format
libname out xport 'path/to/filename.xpt';
data out.mydata;
set sasuser.mydata;
run;

# Then, in R 
library(Hmisc)
mydata <- sasxport.get("path/to/filename.xpt")
# character variables are converted to R factors
# or
library(foreign)
mydata <- read.xport("path/to/filename.xpt")
```



## Exporting Data

- **Text files**:
```{r eval = FALSE}
write.table(mydata, "path/to/filename.txt", sep="\t")  # tab-delimited
write.table(mydata, "path/to/filename.csv", sep=",")   # comma-delimited
write.csv(mydata, "path/to/filename.csv")              # comma-delimited
```

- **Excel spreadsheet**:

```{r eval = FALSE}
library(xlsx)
write.xlsx(mydata, "path/to/filename.xlsx")
```

- **SAS** files:
```{r eval = FALSE}
# write out text datafile and an SAS program to read it
library(foreign)
write.foreign(mydata, datafile = "path/to/filename.txt", 
              codefile = "path/to/filename.sas",  package="SAS")
```

Here are links to more details on data 
[import](http://www.statmethods.net/input/importingdata.html)
and [export](http://www.statmethods.net/input/exportingdata.html). 



## Identifiers 

**Google Convention:** 
  
  * No underscores ( _ ) or hyphens ( - ) in identifiers. 
* **Variable names** all with lower case letters; words separated with dots 
('variable.name'), but 'variableName' is also accepte
* **Function names** have initial capital letters and no dots 
('FunctionName'). Function names should be verbs. Constants are named like 
functions but with an initial k.

```{r, eval = FALSE}
# Good
avg.clicks            # variable name
CalculateAvgClicks    # function name
# Bad
avg_Clicks                                 # variable name
calculate_avg_clicks , calculateAvgClicks  # function name
```

## Identifiers 

**Hadley Convention:**
  
  - Variable and function names should be lowercase. 
- Use an underscore (_) to separate words within a name. 
- Generally, variable names should be nouns and function names should be verbs. 
- Use names that are concise and meaningful. 

```{r eval = FALSE}
# Good
day_one
day_1

# Bad
first_day_of_the_month
DayOne
dayone
djm1
```


## Spacing

- Place spaces around all infix operators (=, +, -, <-, etc.). 

- Always put a space after a comma, and never before (just like in regular English).

- Place a space before left parentheses, except in a function call.

```{r eval = FALSE}
# Good
average <- mean(feet / 12 + inches, na.rm = TRUE)
if (debug) do(x)
plot(x, y)
if (debug) do(x)
diamonds[5, ]

# Bad
average<-mean(feet/12+inches,na.rm=TRUE)
if(debug)do(x)
plot (x, y)
if ( debug ) do(x)  # No spaces around debug
x[1,]               # Needs a space after the comma
x[1 ,]              # Space goes after comma not before
```





```{r}
sessionInfo()
```

