---
title: "Lecture 1: Exercise"
date: September 27th, 2018
output: 
  html_notebook:
    toc: true
    toc_float: true
---

# Exercise 1: Vectors

1. Generate and print a vector of 10 random numbers between 5 and 500.

```{r}
sample(5:500, 10)
```


2. Generate a random vector Z of 1000 letters (from "a" to "z"). Hint: the 
variable `letters` is already defined in R.

```{r}
Z <- sample(letters, 1000, replace = T)
head(Z)
```

3. Print a summary of Z in the form of a frequency table.
```{r}
table(Z)
```


4. Print the list of letters that appear an even number of times in Z.

```{r}
table(Z)[table(Z) %% 2 == 0]
```

# Exercise 2: Matrices

1. Create the following 6 by 6 matrix and store it as variable X.

```{r echo = FALSE}
X <- matrix(1:36, nrow = 6, byrow = FALSE)
X
```

2. Create a matrix Y by adding an independent Gaussian noise (random numbers)
with mean 0 and standard deviation 1 to each entry of X. e.g.

```{r}
Y <- X + matrix(rnorm(36), nrow = 6)
Y
```

3. Find the inverse of Y.

```{r}
invY <- solve(Y)
```


4. Show numerically that the matrix product of Y and its inverse 
is the identity matrix.

```{r}
res <- invY %*% Y
res
```

```{r}
round(res, digits = 6)
```


# Exercise 3: Data fames

1. Create the following data frame and name it “exams”.

```{r, echo = FALSE}
set.seed(123)
(exams <- data.frame(
  student = c("Alice", "Sarah", "Harry", "Ron", "Kate"),
  score = sample(80:100, 5),
  letter = sample(c("A","B"), 5, replace = TRUE),
  late = sample(c(T, F), 5, replace = TRUE)
))
```

2. Compute the mean score for this exam and print it.

```{r}
mean(exams$score)
```

3. Find the student with the highest score and print the corresponding 
row of "exams". Hint: use the function `which.max()`.

```{r}
exams[which.max(exams$score), ]
```

