---
title: "Lecture 2: Exercises"
date: October 5th, 2017
output: html_notebook
---


# Execise 1: Data import/export


Download the data storing New York's Demographic statistics broken down 
by zip code, from the following URL:
"https://cme195.github.io/assets/data/Demographic_Statistics_By_Zip_Code.csv"
to your "~/Desktop" directory:

```{r}
download.file(url = "https://cme195.github.io/assets/data/Demographic_Statistics_By_Zip_Code.csv",
              destfile = "~/Desktop/Demographic_Statistics_By_Zip_Code.csv")
```

Use the following command and check that your file is indeed in the "~/Desktop"
```{r}
list.files("~/Desktop/")
```

Now specify the path to your the file you just downloaded in a variable
called "path2file":

```{r}
# Remember to comment out:
# path2file <- 
```
  
And read-in the text file:
```{r}
# Remember to comment out:
# ny_demographic <- 
```

2. Check the dimensions of your data.


3. Print the variables in your data.


4. Check the content of your data with `str()`, `head()`,
and `summary()` function.


5. Choose only the columns: "COUNT.FEMALE", "COUNT.MALE",  "COUNT.GENDER.UNKNOWN",
and save the data in a file named "Demographic_Statistics_By_Zip_Code_SUBSET.csv"
in the same directory.


# Execise 2: Control Flow 

1. Use a for loop to:
(a) Print the numbers 1 to 15
(b) Print the numbers 5 to 100 that are divisible by 7
(c) Print the numbers 2 to 100 that are divisible by 5 but not by 3
(d) Print for each of the numbers x = 2, . . . 20, all numbers that divide x
(all factors) excluding 1 and x. Hence, for 18, it should print 2 3 6 9.


2. Find all numbers not greater than 10,000 that are divisible by 
5, 7 and 11 and print them.


# Execise 3: Functions

1. Create a function what will return the number
of times a given integer is inside a given vector of integers.
The function should have two arguments one for a vector
and the other for a scalar. Generate a random vector of 100 
integers (in a range 1-20) use the function to count the number
of times 12 is in that vector.


2. Create a function that given an integer will calculate how many divisors 
it has (other than 1 and itself). Make the divisors appear by screen.
Calculate the divisors of 173, 1902 and 3777.


# Execise 4: Apply function family

1. Use `apply` to compute the column means of the `ny_demographic`
data frame. The results for every column (except the first one)
are equal to the average number of participants in the category
indicated by the column name who live in a single jurisdiction/zip code.

2. Using the "COUNT.PARTICIPANTS" column and the rest of the columns, 
compute the percentage of the female, male, pacific islanders, 
hispanic latino, american indian, asian non-hispanic, white non -hispanic, 
black non-hispanic and other ethnicity participants (columns 3-14). 
Append the percentages to the `ny_demographic` dataset.

