---
title: "Lecture 2: Exercises"
date: October 5th, 2017
output: 
  html_notebook:
    toc: true
    toc_float: true
---


# Execise 1: Data import/export

## Part 1.1
Download the file containing data from the 1974 Motor Trend US magazine, 
and comprises fuel consumption and 10 aspects of automobile design and p
erformance for 32 automobiles (1973–74 models) from the following URL:
"https://cme195.github.io/assets/data/Demographic_Statistics_By_Zip_Code.csv"
to your "~/Desktop" directory.

Note that "destfile" argument specifies the path and the name of the file 
you want to download the data to:

```{r}
download.file(url = "https://cme195.github.io/assets/data/mtcars.txt",
              destfile = "~/Desktop/mtcars.txt")
```

Use the following command and check that your file is indeed in your "~/Desktop"
```{r}
list.files("~/Desktop/")
```

a. Specify the path to your the file you just downloaded in a variable
called "path2file".

```{r}
path2file <- "~/Desktop/mtcars.txt"
```
  

Use the `file.show()` command to see the structure of the file.

```{r}
file.show(path2file)
```

Note that the file is tab-delimited. Use suitable argument specifications
to import the data to variable variable "mtcars".

```{r}
mtcars <- read.table(path2file, header = TRUE, sep = "\t")
```


b. Check the content of your data with `str()`, `head()`,
and `summary()` function.

```{r}
head(mtcars)
```

```{r}
str(mtcars)
```

```{r}
summary(mtcars)
```


c. Check the dimensions of your data.

```{r}
dim(mtcars)
```


d. Print the variables in your data.

```{r}
colnames(mtcars)
```



## Part 1.2

Now we will download another dataset on New York's Demographic statistics 
broken down by zip code, from the following URL:
"https://cme195.github.io/assets/data/Demographic_Statistics_By_Zip_Code.csv"
again to your "~/Desktop" directory.

```{r}
download.file(url = "https://cme195.github.io/assets/data/Demographic_Statistics_By_Zip_Code.csv",
              destfile = "~/Desktop/Demographic_Statistics_By_Zip_Code.csv")
```

a. Re-assign the path to the newly downloaded file to the variable "path2file"

```{r}
path2file <- "~/Desktop/Demographic_Statistics_By_Zip_Code.csv"
```

Again, inspect the file with the `file.show()` command.

```{r}
file.show(path2file)
```

Note, that this would not be good to display larger files. An alternative
to see the first few lines of a file is to use a `scan()` function:

```{r}
scan(file = path2file, what = "", sep = "\n", nlines = 10)
```

b. Since, the data is comma-delimited you can use `read.csv()` function
to import the data to variable "demographics", or alternatively use
`read.table()` with suitable arguments:

```{r}
demographics <- read.csv(path2file)
```

c. Use command `head()` to print out the first few rows of the data:

```{r}
head(demographics)
```


c. Choose only the columns: "COUNT.FEMALE", "COUNT.MALE",  "COUNT.GENDER.UNKNOWN",
and save the data in a file named "Demographics_small.csv"
in the same directory.

```{r}
chosenVars <- c("COUNT.FEMALE", "COUNT.MALE", "COUNT.GENDER.UNKNOWN")
filename <- "~/Desktop/Demographics_small.csv"
write.csv(demographics[, chosenVars], file = filename, row.names = FALSE)
```


# Execise 2: Control Flow 

## Part 2.1

Use a for loop to:
a. Print the numbers 1 to 100

```{r}
for (i in 1:15) {
  print(i)
}
```

b. Print the numbers 5 to 100 that are divisible by 7
```{r}
for (i in 5:100) {
  if (i %% 7 == 0) {
    print(i)
  }
}
```

c. Print the numbers 2 to 100 that are divisible by 5 but not by 3.
```{r}
for (i in 2:100) {
  if (i %% 5 == 0 & i %% 3 != 0) {
    print(i)
  }
}
```



## Part 2.2

a. Find all numbers not greater than 10,000 that are divisible by 
5, 7 and 11 and print them.

```{r}
x <- 1
while (x <= 10000) {
  if(all(x %% 5 == 0, x %% 7 == 0, x %% 11 == 0)) {
    print(x)
  }
  x <- x + 1
}
```


b. Print for each of the numbers x = 2, . . . 20, all numbers that divide x
(all factors) excluding 1 and x. Hence, for 18, it should print 2 3 6 9.

```{r}
for (x in 2:20) {
  vec <- rep(NA, x-2)
  for (i in 2:(x-1)) {
    if (x == 2) next
    if (x %% i == 0 ) {
      vec[i-1] <- i 
    }
  }
  cat("Factors of", x, "are:", vec[which(!is.na(vec))], "\n")
}
```

```{r}
# A faster version (inner loop only checks sqrt(x) numbers)
for (x in 2:20) {
  vec <- rep(NA, x-2)
  for (i in 2:ceiling(sqrt(x))) {
    if (x == 2) next
    if (x %% i == 0 ) {
        vec[i - 1] <- i 
        vec[x/i - 1] <- x/i
    }
  }
  cat("Numbers that divide", x, "are:", vec[which(!is.na(vec))], "\n")
}
```


# Execise 3: Functions

# Execise 3: Functions

1. A simple function.
a.Create a function what will return the number
of times a given integer is contained a given vector of integers.
The function should have two arguments one for a vector
and the other for a scalar. 

```{r}
# function returns the number of times a given integer, k, 
# occurs in a given vector of integers, vec.
no_of_integers <- function(vec, k) {
  return(sum(vec == k))
}
```


b. Then, generate a random vector of 100 
integers (in a range 1-20) use the function to count the number
of times the number 12 is in that vector. 

```{r}
(vec <- sample(1:20, 100, replace = TRUE))
cat("Number of times 12 occurs in the vec is:", no_of_integers(vec, 12))
```

2. Write a function that takes in a data.frame as an input, 
prints out the column names, and returns its dimensions.

```{r}
my_function <- function(df) {
  print(colnames(df))
  return(dim(df))
}
```


Extra: Create a function that takes an integer as an input, 
returns how many divisors the integer has (other than 1 and itself),  
and prints them (divisors) all out. 
Then, use the function to find the divisors of 173, 1902 and 3777.

```{r}
# function returns the number divisors of an integer, x,
# (other than 1 and itself)
no_of_divisors <- function(x) {
  vec <- c()
  for (i in 2:ceiling(sqrt(x))) {
    if (x == 2) next
    if (x %% i == 0) {
        vec <- c(vec, i)
        vec <- c(vec, x/i)
    }
  }
  if (length(vec) == 0) return(NULL)
  return(sort(vec))
}
```
 
```{r}
no_of_divisors(173)
no_of_divisors(1902)
no_of_divisors(3777)
```


# Execise 4: Apply function family

1. Use `apply` to compute the column means of the `demographics`
data frame. The results for every column (except the first one)
are equal to the average number of participants in the category
indicated by the column name who live in a single jurisdiction/zip code.

```{r}
apply(demographics, 2, mean)
```


2. Using the "COUNT.PARTICIPANTS" column and the rest of the columns, 
compute the percentage of the female, male, pacific islanders, 
hispanic latino, american indian, asian non-hispanic, white non -hispanic, 
black non-hispanic and other ethnicity participants (columns 3-14). 
Append the percentages to the `demographics` dataset.

```{r}
get_percent <- function(x, total) {
  return(100*x/total)
}
percent.df <- apply(ny_demographic[, 3:14], 2, get_percent, total = demographics[, 2])
colnames(percent.df) <- gsub("COUNT", "PERCENT", colnames(demographics[3:14]))

new.df <- cbind(demographics, percent.df)
head(new.df)
```

