---
title: "Lecture 2: Exercise Answers"
date: October 2nd, 2018
output: 
  html_notebook:
    toc: true
    toc_float: true
---

# Execise 1: Control Flow 

## Part 1.1

Use a for loop to:
a. Print all the letters of the Latin alphabet:

```{r}
for (letter in letters) {
  print(letter)
}
```

b. Print the numbers 10 to 100 that are divisible by 7
```{r}
for (i in 10:100) {
  if (i %% 7 == 0) {
    print(i)
  }
}
```

c. Print the numbers 1 to 100 that are divisible by 5 but not by 3.
```{r}
for (i in 1:100) {
  if (i %% 5 == 0 & i %% 3 != 0) {
    print(i)
  }
}
```



## Part 1.2

a. Find all numbers not greater than 10,000 that are divisible by 
5, 7 and 11 and print them.

```{r}
x <- 1
while (x <= 10000) {
  if(all(x %% 5 == 0, x %% 7 == 0, x %% 11 == 0)) {
    print(x)
  }
  x <- x + 1
}
```


b. Print for each of the numbers x = 2, . . . 20, all numbers that divide x
(all factors) excluding 1 and x. Hence, for 18, it should print 2 3 6 9.

```{r}
for (x in 2:20) {
  vec <- rep(NA, x-2)
  for (i in 2:(x-1)) {
    if (x == 2) next
    if (x %% i == 0 ) {
      vec[i-1] <- i 
    }
  }
  cat("Factors of", x, "are:", vec[which(!is.na(vec))], "\n")
}
```

```{r}
# A faster version (inner loop only checks sqrt(x) numbers)
for (x in 2:20) {
  vec <- rep(NA, x-2)
  for (i in 2:ceiling(sqrt(x))) {
    if (x == 2) next
    if (x %% i == 0 ) {
        vec[i - 1] <- i 
        vec[x/i - 1] <- x/i
    }
  }
  cat("Numbers that divide", x, "are:", vec[which(!is.na(vec))], "\n")
}
```


# Execise 2: Functions

## Part 2.1

a.Create a function what will return the number
of times a given integer is contained a given vector of integers.
The function should have two arguments one for a vector
and the other for a scalar. 

```{r}
# function returns the number of times a given integer, k, 
# occurs in a given vector of integers, vec.
no_of_integers <- function(vec, k) {
  return(sum(vec == k))
}
```


b. Then, generate a random vector of 100 
integers (in a range 1-20) use the function to count the number
of times the number 12 is in that vector. 

```{r}
(vec <- sample(1:20, 100, replace = TRUE))
cat("Number of times 12 occurs in the vec is:", no_of_integers(vec, 12))
```

## Part 2.2 

Write a function that takes in a data.frame as an input, 
prints out the column names, and returns its dimensions.

```{r}
my_function <- function(df) {
  print(colnames(df))
  return(dim(df))
}
```

```{r}
x <- my_function(mtcars)
```
```{r}
x
```

# Execise 3: Apply family functions

## Part 1
Below we print six first rows of the built-in dataset, `mtcars`, 
from the 1974 Motor Trend US magazine, which comprises information
on the fuel consumption and 10 aspects of automobile design 
and performance for 32 selected car models. 

```{r}
head(mtcars)
```

Use `apply()` function to find the standard deviation and the 0.8-quantile  
of each of the automobile characteristic.

```{r}
apply(mtcars, 2, function(x) sd(x))
```


```{r}
apply(mtcars, 2, function(x) { quantile(x, 0.8)})
```

## Part 2

Below is a vector of dates in year 2018.

```{r}
set.seed(1234)
y2018 <- seq(as.Date("2018-01-01", format = "%Y-%m-%d"), 
             as.Date("2018-12-31", format = "%Y-%m-%d"), 
             "days")
length(y2018)
# A random sample of 10 dates from 2018
y2018_sample <- sample(y2018, size = 10)
y2018_sample
```

Use an `apply` family function to return the number of weeks
left from each day in `y2018_sample` to the New Year, 2019/01/01.

```{r}
ny2019 <- as.Date("2019-01-01", format = "%Y-%m-%d")
weeks.to.2019 <- sapply(y2018_sample, function(d) ceiling((ny2019 - d)/7))
names(weeks.to.2019) <- y2018_sample
weeks.to.2019
```

```{r}
sessionInfo()
```

