---
title: "Lecture 2: Exercises"
date: October 5th, 2017
output: html_notebook
---


# Execise 1: Data import/export

## Part 1.1
Download the file containing data from the 1974 Motor Trend US magazine, 
and comprises fuel consumption and 10 aspects of automobile design and p
erformance for 32 automobiles (1973–74 models) from the following URL:
"https://cme195.github.io/assets/data/Demographic_Statistics_By_Zip_Code.csv"
to your "~/Desktop" directory.

Note that "destfile" argument specifies the path and the name of the file 
you want to download the data to:

```{r}
download.file(url = "https://cme195.github.io/assets/data/mtcars.txt",
              destfile = "~/Desktop/mtcars.txt")
```

Use the following command and check that your file is indeed in your "~/Desktop"
```{r}
list.files("~/Desktop/")
```

a. Specify the path to your the file you just downloaded in a variable
called "path2file".

```{r}
# path2file <- 
```
  

Use the `file.show()` command to see the structure of the file.

```{r}
file.show(path2file)
```

Note that the file is tab-delimited. Use suitable argument specifications
to import the data to variable variable "mtcars".

b. Check the content of your data with `str()`, `head()`,
and `summary()` function.

c. Check the dimensions of your data.


d. Print the variables in your data.


## Part 1.2

Now we will download another dataset on New York's Demographic statistics 
broken down by zip code, from the following URL:
"https://cme195.github.io/assets/data/Demographic_Statistics_By_Zip_Code.csv"
again to your "~/Desktop" directory.

```{r}
download.file(url = "https://cme195.github.io/assets/data/Demographic_Statistics_By_Zip_Code.csv",
              destfile = "~/Desktop/Demographic_Statistics_By_Zip_Code.csv")
```

a. Re-assign the path to the newly downloaded file to the variable "path2file"

Again, inspect the file with the `file.show()` command.

Note, that this would not be good to display larger files. An alternative
to see the first few lines of a file is to use a `scan()` function:

```{r}
scan(file = path2file, what = "", sep = "\n", nlines = 10)
```

b. Since, the data is comma-delimited you can use `read.csv()` function
to import the data to variable "demographics", or alternatively use
`read.table()` with suitable arguments:

c. Use command `head()` to print out the first few rows of the data:

d. Choose only the columns: "COUNT.FEMALE", "COUNT.MALE",  "COUNT.GENDER.UNKNOWN",
and save the data in a file named "Demographic_Statistics_By_Zip_Code_SUBSET.csv"
in the same directory.


# Execise 2: Control Flow 

# Execise 2: Control Flow 

## Part 2.1

Use a for loop to:
a. Print the numbers 1 to 100

b. Print the numbers 5 to 100 that are divisible by 7

c. Print the numbers 2 to 100 that are divisible by 5 but not by 3.


## Part 2.2

a. Find all numbers not greater than 10,000 that are divisible by 
5, 7 and 11 and print them.

b. Print for each of the numbers x = 2, . . . 20, all numbers that divide x
(all factors) excluding 1 and x. Hence, for 18, it should print 2 3 6 9.


# Execise 3: Functions

1. A simple function.
a.Create a function what will return the number
of times a given integer is contained a given vector of integers.
The function should have two arguments one for a vector
and the other for a scalar. 

b. Then, generate a random vector of 100 
integers (in a range 1-20) use the function to count the number
of times the number 12 is in that vector. 

2. Write a function that takes in a data.frame as an input, 
prints out the column names, and returns its dimensions.


Extra: Create a function that given an integer will calculate how many 
divisors it has (other than 1 and itself). Make the divisors appear by screen.
Calculate the divisors of 173, 1902 and 3777.


# Execise 4: Apply function family

1. Use `apply` to compute the column means of the `demographics`
data frame. The results for every column (except the first one)
are equal to the average number of participants in the category
indicated by the column name who live in a single jurisdiction/zip code.

2. Using the "COUNT.PARTICIPANTS" column and the rest of the columns, 
compute the percentage of the female, male, pacific islanders, 
hispanic latino, american indian, asian non-hispanic, white non-hispanic, 
black non-hispanic and other ethnicity participants (columns 3-14). 
Append the percentages to the `demographics` dataset.

