---
title: "Lecture 3: Exercises"
date: October 10th, 2017
output: 
  html_notebook:
    toc: true
    toc_float: true
---

```{r global_options, echo = FALSE, include = FALSE}
options(width = 80)
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE, size = "small")
```



# Execise 2: Data manipulation

## Part 1: `dplyr` verbs

Load in the dataset `movies.csv` used in the lecture:

```{r}
url <- "https://raw.githubusercontent.com/Juanets/movie-stats/master/movies.csv"
movies <- tbl_df(read.csv(url))
movies
```


a. Find a subset of the movies produced after 2010. 
Save the subset in 'movies.sub' variable.

b. Keep columns 'name', 'director', 'year', 'country', 'genre', 'budget', 'gross', 'score'
in the 'movies.sub'.

c. Find the profit for each movie in 'movies.sub' as a fraction of its budget.
Convert 'budget' and 'gross' columns million dollar units founded to the first decimal point.
Use `round()` to round numbers.

d. Count the number of movies in 'movies.sub' produced by each genre,
and order them in the descending count order.

e. Now group movies in 'movies.sub' by countries and genre.
Then, count the number of movies in each group and the corresponding 
median fractional profit, the mean and standard deviation of 
the movie score for each group.


## Part 2: Chaining

Using chaining and pipes, for each genre find the three directors the
top mean movie scores received for the movies produced after 2000,
after filtering out the directors with fewer than 5 movies in total.
Hint: Use `top_n()` function to select top n from each group.

Find the movies in your favourite genre by the 5 directors
you just found. These could serve as suggestions for your next movie night!


## Part 3: Merging

Consider the two data-frames below:

```{r}
elections <- data.frame(
  year = rep(c(2016, 2012, 2008, 2004, 2000), each = 2),
  name = c("Hillary Clinton", "Donald Trump", "Barack Obama", "Mitt Romney",
           "Barack Obama", "John McCain", "John Kerry", "George W. Bush",
           "Al Gore", "George W. Bush"),
  party = rep(c("Democratic", "Republican"), 5),
  state = c("NY", "NY", "IL", "MA", "IL", "AZ", "MA", "TX", "TN", "TX")
)
elections
```

```{r}
personal <- data.frame(
  name = c("Hillary Clinton", "Donald Trump", "Barack Obama", "Mitt Romney",
           "John McCain", "George W. Bush"),
  spouse = c("Bill Clinton", "Melania Knauss", "Michelle Robinson", "Ann Davies",
             "Cindy Hensley", "Laura Welch"),
  year = c(1975, 2005, 1992, 1969, 1980, 1977),
  children = c(1, 5, 2, 5, 7, 2)
)
personal
```

a. Merge the two data tables so that you have all the entries from dataset 
'election' but for each candidate you have personal information added.

b. Imagine that you have two data-frames where the columns you want to merge
by do not share the same name like in the example below. Try to merge
the two datasets now, by using appropriate input arguments to `left_join()`.

```{r}
elections <- data.frame(
  year = rep(c(2016, 2012, 2008, 2004, 2000), each = 2),
  candidate = c("Hillary Clinton", "Donald Trump", "Barack Obama", "Mitt Romney",
           "Barack Obama", "John McCain", "John Kerry", "George W. Bush",
           "Al Gore", "George W. Bush"),
  party = rep(c("Democratic", "Republican"), 5),
  state = c("NY", "NY", "IL", "MA", "IL", "AZ", "MA", "TX", "TN", "TX")
)
elections
```

```{r}
personal <- data.frame(
  name = c("Hillary Clinton", "Donald Trump", "Barack Obama", "Mitt Romney",
           "John McCain", "George W. Bush"),
  spouse = c("Bill Clinton", "Melania Knauss", "Michelle Robinson", "Ann Davies",
             "Cindy Hensley", "Laura Welch"),
  year = c(1975, 2005, 1992, 1969, 1980, 1977),
  children = c(1, 5, 2, 5, 7, 2)
)
personal
```




