---
title: "Lecture 5: Exercises with answers"
date: October 17th, 2017
output: 
  html_notebook:
    toc: true
    toc_float: true
---



# Exercise 1: `plotly`

First generate the following (x,y,z) coordinates for your scatter 3D plot:
```{r}
library(plotly)
library(dplyr)

theta <- 2*pi * runif(1000)
phi <- pi * runif(1000)
x <- sin(phi) * cos(theta)
y <- sin(phi) * sin(theta)
z <- cos(phi)
dat <- data.frame(x, y, z)
```

a. Generate a scatter plot with (x, y, z) coordinates, just computed. What do
you observe?



b. Now generate new points as follows. And use `plotly` to create a 3D 
scatter point plot and a separate line plot. Color the points by
their ordering in the data-frame.

```{r}
N <- 10
theta <- rep(2*pi * seq(0, 1, length.out = 100), N)
phi <- rep(pi * seq(0, 1, length.out = N*100))
x <- sin(phi) * cos(theta)
y <- sin(phi) * sin(theta)
z <- cos(phi)
dat <- data.frame(x, y, z)
```


# Exercise 2: heatmaps

How common is your birthday? Download the following data which includes
the records of birthdays on each day of a month in the US between 2000 and 2014. 
The code below aggregates the data across years to count the number of 
birthdays on each day and each the month. Plot a heatmap of `bdays.wide`
to see whether your birthday is among more popular dates. Note that
you should ise settings not to plot dendrograms, as the rows and columns
have their ordering already (according time) and do not need to be clusteres.

```{r}
#install.packages("tidyr")
bdays <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/births/US_births_2000-2014_SSA.csv")
head(bdays)
dim(bdays)
bdays.wide <- bdays %>% 
  group_by(date_of_month, month) %>%
  summarize(births = sum(births)) %>%
  tidyr::spread(key = "date_of_month", value = "births") 
rownames(bdays.wide) <- bdays.wide$month
bdays.wide$month <- NULL
head(bdays.wide)
dim(bdays.wide)
```


# Exercise 3: Plotting maps

a. Use `ggmap` to generate a map of South America, for your conveniences
the location can be set as the bounding box:
`c(-85, -58, -30, 15)`
Choose "stamen" as your source, and "terrain" as your maptype.


b. Use `ggmap` to plot on the same map your house/apartment/dorm and Stanford Main Quad.
Add a path from your house to Stanford (can be in straight line).



