---
title: "Leftover Exercises"
output: html_document
---

# Lecture 2:

## Execise 1: Data import/export

### Part 1.1
Download the file containing data from the 1974 Motor Trend US magazine, 
and comprises fuel consumption and 10 aspects of automobile design and p
erformance for 32 automobiles (1973–74 models) from the following URL:
"https://cme195.github.io/assets/data/mtcars.txt"
to your "~/Desktop" directory.

**NOTE** "destfile" argument specifies the path and the name of the file 
you want to download the data to.
**Windows users need to be careful with path specification**.
For you, the path should look something like this "C:/Users/YourUserName/Desktop".

```{r}
# You should change the "~/Desktop/mtcars.txt" to a path that works for your OS.
download.file(url = "https://cme195.github.io/assets/data/mtcars.txt",
              destfile = "~/Desktop/mtcars.txt")
```

Use the following command and check that your file is indeed in your "~/Desktop"
```{r}
list.files("~/Desktop/")
```

a. Specify the path to your the file you just downloaded in a variable
called "path2file".

```{r}
path2file <- "~/Desktop/mtcars.txt"
```
  

Use the `file.show()` command to see the structure of the file.

```{r}
file.show(path2file)
```

Note that the file is tab-delimited. Use suitable argument specifications
to import the data to variable variable "mtcars".

```{r}
mtcars <- read.table(path2file, header = TRUE, sep = "\t")
```


b. Check the content of your data with `str()`, `head()`,
and `summary()` function.

```{r}
head(mtcars)
```

```{r}
str(mtcars)
```

```{r}
summary(mtcars)
```


c. Check the dimensions of your data.

```{r}
dim(mtcars)
```


d. Print the variables in your data.

```{r}
colnames(mtcars)
```



### Part 1.2

Now we will download another dataset on New York's Demographic statistics 
broken down by zip code, from the following URL:
"https://cme195.github.io/assets/data/Demographic_Statistics_By_Zip_Code.csv"
again to your "~/Desktop" directory.
**Again, Windows users should use path names such as
"C:/Users/YourUserName/Desktop"**.

```{r}
download.file(url = "https://cme195.github.io/assets/data/Demographic_Statistics_By_Zip_Code.csv",
              destfile = "~/Desktop/Demographic_Statistics_By_Zip_Code.csv")
```

a. Re-assign the path to the newly downloaded file to the variable "path2file"

```{r}
path2file <- "~/Desktop/Demographic_Statistics_By_Zip_Code.csv"
```

Again, inspect the file with the `file.show()` command.

```{r}
file.show(path2file)
```

Note, that this would not be good to display larger files. An alternative
to see the first few lines of a file is to use a `scan()` function:

```{r}
scan(file = path2file, what = "", sep = "\n", nlines = 10)
```

b. Since, the data is comma-delimited you can use `read.csv()` function
to import the data to variable "demographics", or alternatively use
`read.table()` with suitable arguments:

```{r}
demographics <- read.csv(path2file)
```

c. Use command `head()` to print out the first few rows of the data:

```{r}
head(demographics)
```


c. Choose only the columns: "COUNT.FEMALE", "COUNT.MALE",  "COUNT.GENDER.UNKNOWN",
and save the data in a file named "Demographics_small.csv"
in the same directory.

```{r}
chosenVars <- c("COUNT.FEMALE", "COUNT.MALE", "COUNT.GENDER.UNKNOWN")
filename <- "~/Desktop/Demographics_small.csv"
write.csv(demographics[, chosenVars], file = filename, row.names = FALSE)
```


# Lecture 3

## Exercise 3: Merging

Consider the two data-frames below:

```{r}
elections <- data.frame(
  year = rep(c(2016, 2012, 2008, 2004, 2000), each = 2),
  name = c("Hillary Clinton", "Donald Trump", "Barack Obama", "Mitt Romney",
           "Barack Obama", "John McCain", "John Kerry", "George W. Bush",
           "Al Gore", "George W. Bush"),
  party = rep(c("Democratic", "Republican"), 5),
  state = c("NY", "NY", "IL", "MA", "IL", "AZ", "MA", "TX", "TN", "TX"),
  stringsAsFactors = FALSE
)
elections
```

```{r}
personal <- data.frame(
  name = c("Hillary Clinton", "Donald Trump", "Barack Obama", "Mitt Romney",
           "John McCain", "George W. Bush"),
  spouse = c("Bill Clinton", "Melania Knauss", "Michelle Robinson", "Ann Davies",
             "Cindy Hensley", "Laura Welch"),
  year = c(1975, 2005, 1992, 1969, 1980, 1977),
  children = c(1, 5, 2, 5, 7, 2), stringsAsFactors = FALSE
)
personal
```

a. Merge the two data tables so that you have all the entries from dataset 
'election' but for each candidate you have personal information added.

Merge 'personal' to 'elections' the two data tables using `left_join()`

```{r}
elections %>% left_join(personal)
```

Observe what happened to the "year" columns.

Note also, that this did not quite get a reasonable result as `left_join()`
by default merges two tables by ALL shared variables, so in this case
"name" AND "year". However, "year" in two data-frames here mean two different 
things, one is the election year and the other is the year married.
What we really want is to **merge by** "name" only.

```{r}
# Note that you need to use QUOTATION marks
elections %>% left_join(personal, by = "name")
```

We can do the reverse operatio and merge 'elections' to 'personal':

```{r}
# Note that this gives you fewer rows:
personal %>% left_join(elections, by = "name")
```

b. Imagine that you have two data-frames where the columns you want to merge
by do not share the same name like in the example below. Try to merge
the two datasets now, by using appropriate input arguments to `left_join()`.

```{r}
elections <- data.frame(
  year = rep(c(2016, 2012, 2008, 2004, 2000), each = 2),
  candidate = c("Hillary Clinton", "Donald Trump", "Barack Obama", "Mitt Romney",
           "Barack Obama", "John McCain", "John Kerry", "George W. Bush",
           "Al Gore", "George W. Bush"),
  party = rep(c("Democratic", "Republican"), 5),
  state = c("NY", "NY", "IL", "MA", "IL", "AZ", "MA", "TX", "TN", "TX")
)
elections
```

```{r}
personal <- data.frame(
  name = c("Hillary Clinton", "Donald Trump", "Barack Obama", "Mitt Romney",
           "John McCain", "George W. Bush"),
  spouse = c("Bill Clinton", "Melania Knauss", "Michelle Robinson", "Ann Davies",
             "Cindy Hensley", "Laura Welch"),
  year = c(1975, 2005, 1992, 1969, 1980, 1977),
  children = c(1, 5, 2, 5, 7, 2)
)
personal
```

```{r}
elections %>% left_join(personal, by = c("candidate" = "name"))
```


