---
title: "Leftover Exercises"
output: html_document
---

# Lecture 2:

## Execise 1: Data import/export

### Part 1.1
Download the file containing data from the 1974 Motor Trend US magazine, 
and comprises fuel consumption and 10 aspects of automobile design and p
erformance for 32 automobiles (1973–74 models) from the following URL:
"https://cme195.github.io/assets/data/mtcars.txt"
to your "~/Desktop" directory.

**NOTE** "destfile" argument specifies the path and the name of the file 
you want to download the data to.
**Windows users need to be careful with path specification**.
For you, the path should look something like this "C:/Users/YourUserName/Desktop".

```{r}
# You should change the "~/Desktop/mtcars.txt" to a path that works for your OS.
download.file(url = "https://cme195.github.io/assets/data/mtcars.txt",
              destfile = "~/Desktop/mtcars.txt")
```

Use the following command and check that your file is indeed in your "~/Desktop"
```{r}
list.files("~/Desktop/")
```

a. Specify the path to your the file you just downloaded in a variable
called "path2file".

```{r}
path2file <- "~/Desktop/mtcars.txt"
```
  

Use the `file.show()` command to see the structure of the file.

```{r}
file.show(path2file)
```

Note that the file is tab-delimited. Use suitable argument specifications
to import the data to variable variable "mtcars".

```{r}
mtcars <- read.table(path2file, header = TRUE, sep = "\t")
```


b. Check the content of your data with `str()`, `head()`,
and `summary()` function.

```{r}
head(mtcars)
```

```{r}
str(mtcars)
```

```{r}
summary(mtcars)
```


c. Check the dimensions of your data.

```{r}
dim(mtcars)
```


d. Print the variables in your data.

```{r}
colnames(mtcars)
```



### Part 1.2

Now we will download another dataset on New York's Demographic statistics 
broken down by zip code, from the following URL:
"https://cme195.github.io/assets/data/Demographic_Statistics_By_Zip_Code.csv"
again to your "~/Desktop" directory.
**Again, Windows users should use path names such as
"C:/Users/YourUserName/Desktop"**.

```{r}
download.file(url = "https://cme195.github.io/assets/data/Demographic_Statistics_By_Zip_Code.csv",
              destfile = "~/Desktop/Demographic_Statistics_By_Zip_Code.csv")
```

a. Re-assign the path to the newly downloaded file to the variable "path2file"

```{r}
path2file <- "~/Desktop/Demographic_Statistics_By_Zip_Code.csv"
```

Again, inspect the file with the `file.show()` command.

```{r}
file.show(path2file)
```

Note, that this would not be good to display larger files. An alternative
to see the first few lines of a file is to use a `scan()` function:

```{r}
scan(file = path2file, what = "", sep = "\n", nlines = 10)
```

b. Since, the data is comma-delimited you can use `read.csv()` function
to import the data to variable "demographics", or alternatively use
`read.table()` with suitable arguments:

```{r}
demographics <- read.csv(path2file)
```

c. Use command `head()` to print out the first few rows of the data:

```{r}
head(demographics)
```


c. Choose only the columns: "COUNT.FEMALE", "COUNT.MALE",  "COUNT.GENDER.UNKNOWN",
and save the data in a file named "Demographics_small.csv"
in the same directory.

```{r}
chosenVars <- c("COUNT.FEMALE", "COUNT.MALE", "COUNT.GENDER.UNKNOWN")
filename <- "~/Desktop/Demographics_small.csv"
write.csv(demographics[, chosenVars], file = filename, row.names = FALSE)
```
